---
title: "Investigating Air Quality with BigRQuery"
author: "Kevin Ushey"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_notebook
---

# Getting Started with BigRQuery

In this R Notebook, we'll use the `bigrquery` package to explore some of the public air quality datasets available on the Google Cloud Platform. These datasets are provided by the US [Environmental Protection Agency](https://www.epa.gov/) and the [OpenAQ](https://openaq.org/) community.

## Loading bigrquery

Note that the `bigrquery` package comes preinstalled on RStudio Server Pro instances launched on Google Cloud; however, if you are running this on a local R instance you can install the package from CRAN. We'll install a number of the packages used in this demo up-front.

```{r}
packages <- c("rmarkdown", "dplyr", "dbplyr", "httpuv", "ggplot2", "bigrquery")
for (package in packages)
  if (!length(find.package(package, quiet = TRUE)))
    install.packages(package)
```

Now that we have these packages installed, let's load them into R up-front.

```{r}
library(bigrquery)
library(dplyr)
library(dbplyr)
```


## Credentials + Project Setup

Normally, the `bigrquery` package will 'just work' -- when you attempt to access BigQuery, OAuth will be used for authentication, and the generated credentials will be cached on your machine. For the purposes of my demo, I wanted to skip the OAuth step here, so I have the credentials already cached on the system as per the directions in <https://developers.google.com/identity/protocols/application-default-credentials?hl=en_US> -- but you should be able to omit this step if you prefer.


```{r}
# set project name
project <- "perfect-lantern-179621"

# set up credentials
credentials <- readr::read_file("credentials.json")
set_service_token(credentials)
```


## First Steps

First, let's try running some raw SQL with `query_exec()` to confirm that we've got our environment set up as expected. We'll try looking at the first 20 rows of the `air_quality_annual_summary` data set, from the `epa_historical_air_quality` database. (See <https://bigquery.cloud.google.com/dataset/bigquery-public-data:epa_historical_air_quality> for more details on this dataset.)

```{r}
# generate our SQL query (as an R string)
sql <- "
  SELECT
    *
  FROM 
    [bigquery-public-data:epa_historical_air_quality.air_quality_annual_summary]
  LIMIT
    20
"

# run it with 'query_exec()'
query_exec(sql, project = project)
```

`query_exec()` is handy when you want to execute a one-off query, or a set of SQL that you've either defined inline or perhaps as part of a separate script (e.g. a `.sql` file).


## Using DBI

The `bigrquery` package also provides a DBI driver, which makes it possible to talk to Google's BigQuery using the `DBI` generics you might already be familiar with.

```{r}
library(DBI)

# construct a DBI connection using bigrquery's DBI driver to the
# 'epa_historical_air_quality' database
conn <- DBI::dbConnect(
  bigrquery::dbi_driver(),
  project = "bigquery-public-data",
  dataset = "epa_historical_air_quality",
  billing = project
)

# list the tables available in this database
DBI::dbListTables(conn)
```

Great! Let's try running the same query we used above. Now that we've established a connection directly to the `epa_historical_air_quality` database, it's no longer necessary to specifiy the full `[database:table]` name in our SQL statement -- we can reference the `air_quality_annual_summary` table directly.

```{r}
# run a SQL statement directly
sql <- "SELECT * FROM air_quality_annual_summary LIMIT 20"
DBI::dbGetQuery(conn, sql)
```

In fact, once we have this DBI connection available, we can write a SQL code block directly to construct and execute a query.

```{sql, connection="conn"}
SELECT * FROM air_quality_annual_summary LIMIT 20
```


## Using dplyr

The `dplyr` package provides a flexible grammar for data manipulation. It has three main goals:

- Identify the most important data manipulation verbs and make them easy to use from R.

- Provide blazing fast performance for in-memory data by writing key pieces in C++ (using Rcpp)

- Use the same interface to work with data no matter where it's stored, whether in a data frame, a data table or database.

This implies that the same code you might use to interact with an in-memory R data.frame will look very similar to the code you might write to access a table stored in a SQL database. Let's get `dplyr` talking to Google BigQuery:

```{r}
# construct a 'src': this is the connection object that dplyr uses
# to communicate with Google BigQuery
src <- bigrquery::src_bigquery(
  project = "bigquery-public-data",
  dataset = "epa_historical_air_quality",
  billing = project
)

# generate a reference to the 'air_quality_annual_summary' table
aqas <- tbl(src, "air_quality_annual_summary")

head(aqas, n = 20)
```

Awesome! Now, let's try using `dplyr` to answer the question, "what was the average air quality in each year, by state?". In the context of SQL databases, you can think of `dplyr` as a generator for SQL select statements.

```{r}
# construct our query using the dplyr interface
query <- aqas %>%
  select(state_name, year, units_of_measure, arithmetic_mean) %>%
  group_by(state_name, year) %>%
  summarize(avg = mean(arithmetic_mean))

# inspect the SQL actually generated here
sql_render(query)
```

Now, let's try running that query:

```{r}
# use 'collect()' to execute the query, and retrieve
# the results as an R data.frame
quality <- collect(query)
quality
```


## Interlude

You might be a bit confused by the `%>%` operator above. This is a user-defined operator provided by the `magrittr` package (whose name is inspired by René Magritte's "Ceci n'est pas une pipe"). It's similar to the F# pipe-forward operator, and its main utility is unwrapping nested function calls and avoiding temporary variables. Compare:

```{r}
# without the pipe operator, you might write
analyze(subset(prepare(dataset)))

# with the pipe, you could unfold this as
data %>% prepare() %>% subset() %>% analyze()
```

It helps to read the `%>%` operator as the "and then" operator in the context of `dplyr`. We can use it to great effect with the `dplyr` package.


## Back on Track

Awesome! Now, let's switch gears a bit. The annual summary dataset aggregates a lot of disparate air quality measures, which means normalizing the various data is a bit out of scope for this talk. Let's look at the `pm25_frm_daily_summary` dataset, which (roughly speaking) measures the density of particulate matter of 25 microns or less in the air. (A description of the table is available at <https://bigquery.cloud.google.com/table/bigquery-public-data:epa_historical_air_quality.pm25_frm_daily_summary>.)

```{r}
pm25 <- tbl(src, "pm25_frm_daily_summary")
head(pm25, n = 20)
```

Let's look at the yearly average PM25.

```{r}
avg <- pm25 %>%
  select(state_name, arithmetic_mean, date_local) %>%
  mutate(year = extract(year %from% date_local)) %>%
  group_by(state_name, year) %>%
  summarize(mean = mean(arithmetic_mean)) %>%
  arrange(state_name, year) %>%
  collect()

head(avg, n = 20)
```

And let's take a look at the trends over time. We'll use the `ggplot2` package to visualize.

```{r}
library(ggplot2)

plot_average <- function(data) {
  ggplot(data, aes(x = year, y = mean)) +
    geom_line() +
    facet_wrap(~ state_name) +
    labs(x = "Year", y = expression(Average ~ PM[25] ~ Concentration)) +
    ggtitle(expression(Average ~ Historical ~ PM[25] ~ Concentration)) +
    theme(
      axis.text.x = element_text(size = 6, angle = 45),
      axis.text.y = element_text(size = 6)
    ) 
}

plot_average(avg)
```

Looks like there was a big spike in California's pollution in the early 2000. Let's remove that data just to make the trend for other states more clear:

```{r}
avg %>%
  filter(year > 2000) %>%
  plot_average()
```

Cool! Let's do one last cool thing -- we'll use the `leaflet` package to visualize the last hour of Air Quality data, as provided by [OpenAQ](https://openaq.org/) and made available in the `global_air_quality` dataset.

```{r}
src <- bigrquery::src_bigquery(
  project = "bigquery-public-data",
  dataset = "openaq",
  billing = project
)

tbl <- tbl(src, "global_air_quality")
head(tbl, n = 20)
```

```{r}
library(leaflet)

# define our query and collect the data into R
pm25 <- tbl %>%
  filter(pollutant == "pm25", country == "US") %>%
  select(location, city, country, value, latitude, longitude, timestamp) %>%
  arrange(value) %>%
  collect()

# define a palette used for coloring points on our map
# we'll color based on the measured pm10 (higher is darker orange)
pal <- colorNumeric("Oranges", domain = pm25$value)

# format used for labels
fmt <- "[%s] %s, %s, %s: (%s)"

# construct our map
leaflet(pm25) %>%
  addTiles() %>%
  addCircleMarkers(
    lat = ~ latitude,
    lng = ~ longitude,
    color = ~ pal(value),
    label = ~ sprintf(fmt, timestamp, location, city, country, value),
    stroke = FALSE,
    fillOpacity = 0.6
  ) %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = ~ value,
    title = "Latest Measured PM25"
  )
```

---

Thanks for attending! If you'd like to learn more about RStudio, please visit us at <https://www.rstudio.com>.

The materials for this demo are available at <https://github.com/kevinushey/2017-google-cloud-bigquery>.






